---
title: 'Phylogenetic Tree Annotation'
---

```{r, echo=FALSE, message=FALSE, eval=TRUE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
options(digits=3)
options(max.print=200)
library(ggtree)

```



This lesson introduces basic annotation utilies for common tasks such as highlight and labelling clades.

## Label Clade

```{r}
library(ggtree)
set.seed(2015-12-21)
tree <- rtree(30)
p <- ggtree(tree) + xlim(NA, 6)

p + geom_cladelabel(node=45, label="test label") +
    geom_cladelabel(node=34, label="another clade")
```

We can use `align=TRUE` to align the clade label and use `offset` to adjust the position.

```{r}
p + geom_cladelabel(node=45, label="test label", align=T, color='red') +
    geom_cladelabel(node=34, label="another clade", align=T, color='blue')
```

### Exercise 1

Look at the help for `?geom_cladelabel` to learn the usages of other parameters.

1. rotate the text 'another clade` with 45 degree.
2. increase the sizes of bar and label for the 'test label'
3. increase relative positon of bar and label


```{r eval=F, echo=F}
p + geom_cladelabel(node=45, label="test label", align=T) +
    geom_cladelabel(node=34, label="another clade", align=T, angle=45)

p + geom_cladelabel(node=45, label="test label", align=T, barsize=1.5, fontsize=8) +
    geom_cladelabel(node=34, label="another clade", align=T, angle=45)

p + geom_cladelabel(node=45, label="test label", align=T, barsize=1.5, fontsize=8, offset.text=.3) +
    geom_cladelabel(node=34, label="another clade", align=T, angle=45)
```


## Highlight clades

**ggtree** implements `geom_hilight` to add rectangle to highlight selected clade.


```{r}
ggtree(tree) + geom_hilight(node=34, fill="steelblue", alpha=.6) +
    geom_hilight(node=45, fill="darkgreen", alpha=.6)
```

`geom_hilight` supports rectangular and circular layouts. For unrooted layout, **ggtree** provides `geom_hilight_encircle` layer.

```{r}
ggtree(tree, layout = "unrooted") + geom_hilight_encircle(node=34, fill='steelblue') +
  geom_hilight_encircle(node=45, fill="darkgreen")
```

## Exercise 2

1. Create a circular layout tree and highlight the selected clades.

```{r eval=F, echo=F}
ggtree(tree, layout='circular') + geom_hilight(node=34, fill="steelblue", alpha=.6) +
    geom_hilight(node=45, fill="darkgreen", alpha=.6)
```



## Uncertainty of evolutionary inference

**ggtree** provides `geom_range` to display uncertainty of evolutionary inference.

This example use a BEAST tree, which was imported by [treeio](r-treeio.html). Data that stored in the tree object or mapped to the tree from [external data](r-treeio.html) can be used to annotate the tree directly.


```{r fig.width=6, fig.height=8}
beast_tree <- read.beast("data/MCC_FluA_H3.tree")
ggtree(beast_tree) + geom_range(range = 'height_0.95_HPD', branch.length='height', color='red', alpha=.6, size=2)
```

### Exercise 3

1. create a heat tree by using substitution rate (`rate` in `beast_tree`) to color branches.

## Visualize tree with associated matrix (as a heatmap)

Visualizing [tree with external data](r-ggtree-data.html) will mainly be introduced in next lesson. As visualizing heatmap and multiple sequence alignment with the tree is more frequently used, **ggtree** provides `gheatmap` and `msaplot` function to these tasks.



